\textbf{COMP0085 Summative Assignment}

### 2022-12-25 submission.tex
fix free energy
update to gp question

We want the posterior mean and covariance over $a$ and $b$.
Defining a weight vector $\textbf{w}$:

\[\textbf{w} = \begin{bmatrix}
                 a \\
                 b
         \end{bmatrix}\]

Our distribution for $\textbf{w}$:

\[P(\textbf{w}) = \mathcal{N} \left(
\begin{bmatrix}
                 \mu_a \\
                 \mu_b
         \end{bmatrix} ,
\begin{bmatrix}
                 \sigma_a^2 & 0 \\
                 0 & \sigma_b^2
         \end{bmatrix}
\right)
  = \mathcal{N}(\mu_{\textbf{w}}, \Sigma_{\textbf{w}})
\]

Moreover, for our data $\mathcal{D} = \{\textbf{X}, \textbf{Y}\}$:

\[P(\mathcal{D} | \textbf{w}) = \mathcal{N} \left( \textbf{Y} - \textbf{w}^T \textbf{X}, \sigma^2 \textbf{I}
\right)
\]

@@

Knowing:

\[P(\textbf{w} | \mathcal{D}) \propto P(\mathcal{D} | \textbf{w}) P(\textbf{w})\]

we can substitute the above distributions:

\[P(\textbf{w} | \mathcal{D}) \propto
  \exp \left(\frac{-1}{2 \sigma^2} \left( \textbf{Y} - \textbf{w}^T \textbf{X}\right) \left( \textbf{Y} - \textbf{w}^T \textbf{X}\right)^T \right)
\exp \left(\frac{-1}{2} \left( \textbf{w} -\mu_{\textbf{w}}\right)^T \Sigma_{\textbf{w}}^{-1} \left( \textbf{w} -\mu_{\textbf{w}}\right)\right)
\]

expanding:

\[\log P(\textbf{w} | \mathcal{D}) \propto
  \frac{-1}{2} \left( \frac{\textbf{Y} \textbf{Y}^T}{\sigma^2} - 2\textbf{w}^T \frac{\textbf{X}\textbf{Y}^T}{\sigma^2} + \textbf{w}^T \frac{\textbf{X} \textbf{X}^T}{\sigma^2} \textbf{w} +  \textbf{w}^T \Sigma_{\textbf{w}}^{-1}  \textbf{w} - 2\textbf{w}^T \Sigma_{\textbf{w}}^{-1}\mu_{\textbf{w}} + \mu_{\textbf{w}}^T\Sigma_{\textbf{w}}^{-1}\mu_{\textbf{w}} \right)
\]

collecting $\textbf{w}$ terms:

\[\log P(\textbf{w} | \mathcal{D}) \propto
  \frac{-1}{2} \left( \textbf{w}^T \left(\frac{\textbf{X} \textbf{X}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \right)  \textbf{w} - 2\textbf{w}^T \left(\frac{\textbf{X}\textbf{Y}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \mu_{\textbf{w}}\right)  \right)
\]

Knowing that the posterior $P(\textbf{w} | \mathcal{D})$ will be Gaussian with mean $\bar{\mu}_w$ and covariance $\bar{\Sigma}_w$, we can see that expanding the exponent component would have the form:

\[\left( \textbf{w} - \bar{\mu}_w \right)^T \bar{\Sigma}_w^{-1} \left( \textbf{w} - \bar{\mu}_w \right) = \textbf{w}^T \bar{\Sigma}_w^{-1} \textbf{w} -2 \textbf{w}^T \bar{\Sigma}_w^{-1} \bar{\mu}_w + \bar{\mu}_w^T \bar{\Sigma}_w^{-1} \bar{\mu}_w\]

Thus we can identify the posterior covariance:

\[\bar{\Sigma}_w = \left(\frac{\textbf{X} \textbf{X}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \right)^{-1}\]

and the posterior mean:

\[\bar{\mu}_w = \bar{\Sigma}_w \left(\frac{\textbf{X}\textbf{Y}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \mu_{\textbf{w}}\right) \]

\newpage
The Python code:
\lstinputlisting[language=Python]{src/solutions/q2.py}

@@

\[ \log P(\textbf{x} | \textbf{s}, \theta)  = -\frac{d}{2} \log (2 \pi) - \log(\sigma)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K} s_i \mu_i   + \sum_{i=1}^{K} \sum_{i=1}^{K} s_i s_j \mu_i^T\mu_j \right) \]

@@

\[
\langle \log P(\textbf{x} | \textbf{s}, \theta)\rangle_{q(\textbf{s})} \\
= -\frac{d}{2} \log (2 \pi) - \log(\sigma)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K} \langle s_i \rangle_{q_{i} (s_i)} \mu_i   + \sum_{i=1}^{K} \sum_{j=1}^{K} \langle s_i s_j \rangle_{q_{i} (s_i) q_{j} (s_j)} \mu_i^T\mu_j \right)\]

@@

\[
\langle \log P(\textbf{x} | \textbf{s}, \theta)\rangle_{q(\textbf{s})} \\
= -\frac{d}{2} \log (2 \pi) - \log(\sigma)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K}  \lambda_i  \mu_i   + \sum_{i=1}^{K} \sum_{j=1, j\neq i}^{K}  \lambda_i \lambda_j \mu_i^T\mu_j + \sum_{i=1}^{K}  \lambda_i \mu_i^T\mu_i \right)\]

@@

\[
\begin{array}{l}
\tab \mathcal{F}(q, \theta) = \\
\tab \tab \frac{-d}{2} \log (2 \pi) - \log(\sigma)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K}  \lambda_i  \mu_i   + \sum_{i=1}^{K} \sum_{j=1, j\neq i}^{K}  \lambda_i \lambda_j \mu_i^T\mu_j + \sum_{i=1}^{K}  \lambda_i \mu_i^T\mu_i \right) \\
\tab \tab + \sum_{i=1}^{K} \lambda_i \log \pi_i + (1-\lambda_i) \log (1-\pi_i)}\\
\tab \tab - \sum_{i=1}^K \lambda_i \log \lambda_i + (1-\lambda_i) \log (1-\lambda_i)
\end{array}
\]

>>>>

We want the posterior mean and covariance over $a$ and $b$.
Defining a weight vector $\textbf{w}$:
\[\textbf{w} = \begin{bmatrix}
                 a \\
                 b
         \end{bmatrix}\]

Our distribution for $\textbf{w}$:
\[P(\textbf{w}) = \mathcal{N} \left(
\begin{bmatrix}
                 \mu_a \\
                 \mu_b
         \end{bmatrix} ,
\begin{bmatrix}
                 \sigma_a^2 & 0 \\
                 0 & \sigma_b^2
         \end{bmatrix}
\right)
  = \mathcal{N}(\mu_{\textbf{w}}, \Sigma_{\textbf{w}})
\]

Moreover, for our data $\mathcal{D} = \{\textbf{X}, \textbf{Y}\}$:
\[P(\mathcal{D} | \textbf{w}) = \mathcal{N} \left( \textbf{Y} - \textbf{w}^T \textbf{X}, \sigma^2 \textbf{I}
\right)
\]

@@

Knowing:
\[P(\textbf{w} | \mathcal{D}) \propto P(\mathcal{D} | \textbf{w}) P(\textbf{w})\]

we can substitute the above distributions:
\[P(\textbf{w} | \mathcal{D}) \propto
  \exp \left(\frac{-1}{2 \sigma^2} \left( \textbf{Y} - \textbf{w}^T \textbf{X}\right) \left( \textbf{Y} - \textbf{w}^T \textbf{X}\right)^T \right)
\exp \left(\frac{-1}{2} \left( \textbf{w} -\mu_{\textbf{w}}\right)^T \Sigma_{\textbf{w}}^{-1} \left( \textbf{w} -\mu_{\textbf{w}}\right)\right)
\]

expanding:
\[\log P(\textbf{w} | \mathcal{D}) \propto
  \frac{-1}{2} \left( \frac{\textbf{Y} \textbf{Y}^T}{\sigma^2} - 2\textbf{w}^T \frac{\textbf{X}\textbf{Y}^T}{\sigma^2} + \textbf{w}^T \frac{\textbf{X} \textbf{X}^T}{\sigma^2} \textbf{w} +  \textbf{w}^T \Sigma_{\textbf{w}}^{-1}  \textbf{w} - 2\textbf{w}^T \Sigma_{\textbf{w}}^{-1}\mu_{\textbf{w}} + \mu_{\textbf{w}}^T\Sigma_{\textbf{w}}^{-1}\mu_{\textbf{w}} \right)
\]

collecting $\textbf{w}$ terms:
\[\log P(\textbf{w} | \mathcal{D}) \propto
  \frac{-1}{2} \left( \textbf{w}^T \left(\frac{\textbf{X} \textbf{X}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \right)  \textbf{w} - 2\textbf{w}^T \left(\frac{\textbf{X}\textbf{Y}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \mu_{\textbf{w}}\right)  \right)
\]

Knowing that the posterior $P(\textbf{w} | \mathcal{D})$ will be Gaussian with mean $\bar{\mu}_w$ and covariance $\bar{\Sigma}_w$, we can see that expanding the exponent component would have the form:
\[\left( \textbf{w} - \bar{\mu}_w \right)^T \bar{\Sigma}_w^{-1} \left( \textbf{w} - \bar{\mu}_w \right) = \textbf{w}^T \bar{\Sigma}_w^{-1} \textbf{w} -2 \textbf{w}^T \bar{\Sigma}_w^{-1} \bar{\mu}_w + \bar{\mu}_w^T \bar{\Sigma}_w^{-1} \bar{\mu}_w\]

Thus we can identify the posterior covariance:
\[\bar{\Sigma}_w = \left(\frac{\textbf{X} \textbf{X}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \right)^{-1}\]

and the posterior mean:
\[\bar{\mu}_w = \bar{\Sigma}_w \left(\frac{\textbf{X}\textbf{Y}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \mu_{\textbf{w}}\right) \]

\newpage
Computing the posterior mean and covariance over $a$ and $b$ given by the $CO_2$ data:

\begin{figure}[h]
\centering
\begin{minipage}{.5\textwidth}
  \centering
\includegraphics[scale = 0.5]{outputs/q2/a-mean}
\caption{The Posterior Mean}
\label{fig:fig2-a-mean}
\end{minipage}%
\begin{minipage}{.5\textwidth}
  \centering
\includegraphics[scale = 0.5]{outputs/q2/a-covariance}
\caption{t=The Posterior Covariance}
\label{fig:fig2-a-covariance}
\end{minipage}
\end{figure}

Plotting the residuals:

\begin{figure}[h]
\centering
\includegraphics[scale=0.5]{outputs/q2/b-residuals-timeseries}
\caption{$g_{obs}(t)$}
\label{fig:b-residuals-timeseries}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[scale=0.5]{outputs/q2/b-residuals-density-estimation}
\caption{Density Estimation of Residuals vs $e(t) \sim \mathcal{N}(0, 1)$}
\label{fig:b-residuals-density-estimation}
\end{figure}

We can see that the residuals do not perfectly conform to our prior over $e(t) \sim \mathcal{N}(0, 1)$.
The density estimation shows that a mean of zero is a reasonable prior belief however the data does not seem to exhibit unit variance.

\newpage
\subsection*{(c \& d)}

We are considering the kernel:

\[k(s, t) = \theta^2 \left( \exp\left( - \frac{2\sin^2(\pi(s-t)/\tau)}{\sigma^2}\right) + \phi^2 \exp\left( - \frac{(s-t)^2}{2\eta^2}\right)\right) + \zeta^2 \delta_{s=t}\]

We can make qualitative observations this kernel by visualising the covariance (gram) matrix:

\begin{figure}[h]
\centering
\includegraphics[scale=0.5]{outputs/q2/c-gram-matrix}
\caption{Gram Matrix}
\label{fig:c-gram-matrix}
\end{figure}

We can observe a striped pattern which indicate higher covariance at regular intervals.
This can be attributed to the sinusoidal term in the kernel and encourages sinusoidal functions.
Additionally, we can see that covariance values also decay as they are further away from the diagonal.
This can be attributed to the exponential term in the kernel, encouraging points closer in time to be more correlated and vice versa.
From our $CO_2$ data, we would want a class of functions which exhibit both of these behaviours as the data looks sinusoidal (seasonal with respect to each year) and correlations locally.

We can also visualise some samples from a Gaussian Process with the same covariance matrix and zero mean.
This verifies our observations about the covariance matrix.

\begin{figure}[h]
\centering
\includegraphics[scale=0.5]{outputs/q2/c-samples}
\caption{Samples from a zero mean GP with the provided covariance kernel}
\label{fig:c-samples}
\end{figure}

\newpage
More specifically, we can see how changing each hyper-parameter will affect the characteristics of the function.

\begin{figure}[h]
\centering
\begin{minipage}{.5\textwidth}
  \centering
\includegraphics[scale = 0.35]{outputs/q2/c-parameter-grams}
\caption{Covariances for different parameters}
\label{fig:fig2-c-parameter-grams}
\end{minipage}%
\begin{minipage}{.5\textwidth}
  \centering
\includegraphics[scale = 0.35]{outputs/q2/c-parameter-samples}
\caption{Samples for different parameters}
\label{fig:fig2-c-parameter-samples}
\end{minipage}
\end{figure}

\begin{itemize}
\item [$\theta$:] As $\theta$ increases, we see more pronounced periodic behavior in the sample function.
    The covariance matrix shows how increasing $\theta$ visually reveals the striped periodic component.
        This is expected because it is the parameter that adjusts the weight of $\exp\left( - \frac{2\sin^2(\pi(s-t)/\tau)}{\sigma^2}\right)$.
\item [$\sigma$:] As $\sigma$ increases, we see smoother periodic behaviour in the sample function.
    The covariance matrix shows how increasing $\sigma$ will increase covariance values in the off-diagonals.
    This is expected because it adjusts the lengthscale of the periodic portion of the kernel.
\item [$\phi$:] As $\phi$ increases, we see less smooth behaviour in the sample function.
    The covariance matrix shows how increasing $\sigma$ will increase covariance values in the off-diagonals.
    This is expected because it adjusts the lengthscale of the periodic portion of the kernel.
\item [$\eta$:]
\item [$\tau$:]
\item [$\zeta$:]
\end{itemize}
\newpage
\subsection*{(e)}

\subsection*{(f)}
\begin{figure}[h]
\centering
\includegraphics[scale=0.5]{outputs/q2/f-extrapolation}
\caption{Extrapolation of $CO_2$ concentration levels}
\label{fig:f-extrapolation}
\end{figure}

\subsection*{(g)}

\newpage
The Python code for Bayesian Linear Regression:
\lstinputlisting[language=Python]{src/models/bayesian_linear_regression.py}

\newpage
The Python code for kernels:
\lstinputlisting[language=Python]{src/models/kernels.py}

\newpage
The Python code for Gaussian Process Regression:
\lstinputlisting[language=Python]{src/models/gaussian_process_regression.py}

\newpage
The rest of the Python code for question 2:
\lstinputlisting[language=Python]{src/solutions/q2.py}

@@

\[ \log P(\textbf{x} | \textbf{s}, \theta)  = -\frac{d}{2} \log (2 \pi \sigma^2)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K} s_i \mu_i   + \sum_{i=1}^{K} \sum_{i=1}^{K} s_i s_j \mu_i^T\mu_j \right) \]

@@

\[
\langle \log P(\textbf{x} | \textbf{s}, \theta)\rangle_{q(\textbf{s})} \\
= -\frac{d}{2} \log (2 \pi \sigma^2)   -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K} \langle s_i \rangle_{q_{i} (s_i)} \mu_i   + \sum_{i=1}^{K} \sum_{j=1}^{K} \langle s_i s_j \rangle_{q_{i} (s_i) q_{j} (s_j)} \mu_i^T\mu_j \right)\]

@@

\[
\langle \log P(\textbf{x} | \textbf{s}, \theta)\rangle_{q(\textbf{s})} \\
= -\frac{d}{2} \log (2 \pi \sigma^2)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K}  \lambda_i  \mu_i   + \sum_{i=1}^{K} \sum_{j=1, j\neq i}^{K}  \lambda_i \lambda_j \mu_i^T\mu_j + \sum_{i=1}^{K}  \lambda_i \mu_i^T\mu_i \right)\]

@@

\[
\begin{array}{l}
\tab \mathcal{F}(q, \theta) = \\
\tab \tab \frac{-d}{2} \log (2 \pi \sigma^2)   -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K}  \lambda_i  \mu_i   + \sum_{i=1}^{K} \sum_{j=1, j\neq i}^{K}  \lambda_i \lambda_j \mu_i^T\mu_j + \sum_{i=1}^{K}  \lambda_i \mu_i^T\mu_i \right) \\
\tab \tab + \sum_{i=1}^{K} \lambda_i \log \pi_i + (1-\lambda_i) \log (1-\pi_i)}\\
\tab \tab - \sum_{i=1}^K \lambda_i \log \lambda_i + (1-\lambda_i) \log (1-\lambda_i)
\end{array}
\]
